String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-03 2023-1-19/sections/cefa.tex
operations for NFA

The semantics of CEFA is define as follows. Let $\aut = (R, Q, \Sigma, \delta, q_0, F, \alpha)$ be a CEFA. 
A \emph{configuration} of $\aut$ is a pair $(q, \vec{v})$ where $q \in Q$ and $\vec{v}$ is a vector denoting the values of registers.  
The \emph{initial configuration} of $\aut$ is $(q_0, \vec{0})$, where the value of each register is zero. 
A \emph{run} of $\aut$ on a string $w = a_1 \cdots a_n$ is a sequence $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ and $q_{i-1} \xrightarrow[\myvec{v_i}]{a_i} q_i$ for each $i \in [n]$. A run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is \emph{accepting} if $q_n \in F$. The \emph{cost} of an accepting run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is defined as $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}$.Note here we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed.  We use $\myvec{v'} \in \aut(w)$ to denote the fact that there is an accepting run of $\aut$ on $w$ whose cost is $\myvec{v'}$. 
Finally, we define the semantics of a CEFA $\aut$ as $\Lang(\aut) = \{(w, \myvec{v'}) \mid  \myvec{v'} \in \aut(w)\}$.

@@

%Finally, we define the product of two CEFA. 
Let $\aut_1 = (R_1, Q_1, \Sigma, \delta_1, q_{1,0}, F_1, \alpha_1)$ and  $\aut_2 = (R_2, Q_2, \Sigma, \delta_2, q_{2,0}, F_2, \alpha_2)$ be two CEFA that share the alphabet. Moreover, suppose $R_1 \cap R_2 = \emptyset$. We define the \emph{product} of $\aut_1$ and $\aut_2$, denoted by $\aut_1 \times \aut_2 = (R', Q', \Sigma, \delta', q'_0, F', \alpha')$, as follows. 
\begin{itemize}
\item $R' = R_1 \cup R_2$, $Q' = Q_1 \times Q_2$, $q'_0 = (q_{1,0}, q_{2,0})$, $F' = F_1 \times F_2$,
%
%\item $\alpha'((q_1, q_2)) = \alpha_1(q_1) \wedge \alpha_2(q_2)$ for each $(q_1, q_2) \in F'$, 
%
\item $\delta'$ comprises the tuples $((q_1, q_2), a, (q'_1, q'_2), \myvec{v})$ such that $(q_1,a, q'_1, \prj_{R_1}(\myvec{v})) \in \delta_1$ and $(q_2,a, q'_2, \prj_{R_2}(\myvec{v})) \in \delta_2$, where $\myvec{v} \in \Int^{R'}$, $\prj_{R_1}(\myvec{v}))$ (resp. $\prj_{R_2}(\myvec{v}))$) is the restriction of $\myvec{v}$ to the domain $R_1$ (resp. $R_2$). 
\end{itemize}
Roughly speaking, $\aut_1 \times \aut_2$ accepts the words that are accepted by both $\aut_1$ and $\aut_2$ and outputs the costs of both $\aut_1$ and $\aut_2$.  
%the intersection of $\Lang(\aut_1)$ and $\Lang(\aut_2)$.

Let $((\aut_1,\cdots, \aut_n), \varphi)$ be a tuple such that $\aut_1 = (R_1, Q_1, \Sigma, \delta_1, q_{1,0}, F_1)$, $\cdots$, $\aut_n = (R_n, Q_n, \Sigma, \delta_n, q_{n, 0}, F_n)$ are CEFA, and $\varphi(R_1, \cdots, R_n)$ is an LIA formula. Then the \emph{relation} defined by $((\aut_1, \cdots, \aut_n), \varphi)$, denoted by $\srel((\aut_1, \cdots, \aut_n), \varphi)$, is $\{(w, \myvec{v'_1}, \cdots, \myvec{v'_n}) \mid \forall i \in [n].\ (w, \myvec{v'_i}) \in \Lang(\aut_i), \varphi(\myvec{v'_1}/R_1, \cdots, \myvec{v'_n}/R_n) \mbox{ is true}\}$.

>>>>

The semantics of CEFA is define as follows. Let $\aut = (R, Q, \Sigma, \delta, q_0, F, \alpha)$ be a CEFA. 
A \emph{configuration} of $\aut$ is a pair $(q, \vec{v})$ where $q \in Q$ and $\vec{v}$ is a vector denoting the values of registers.  
The \emph{initial configuration} of $\aut$ is $(q_0, \vec{0})$, where the value of each register is zero. 
A \emph{run} of $\aut$ on a string $w = a_1 \cdots a_n$ is a sequence $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ and $q_{i-1} \xrightarrow[\myvec{v_i}]{a_i} q_i$ for each $i \in [n]$. A run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is \emph{accepting} if $q_n \in F$. The \emph{cost} of an accepting run $q_0 \xrightarrow[\myvec{v_1}]{a_1} q_1 \cdots q_{n-1}\xrightarrow[\myvec{v_n}]{a_n} q_n$ is defined as $\myvec{v'} = \sum \limits_{j \in [n]} \myvec{v_j}$. Note here we assume that the initial values of all registers are zero and $\sum \limits_{j \in [n]} \myvec{v_j}$ is the vector of register values after all the transitions in the run are executed.  We use $\myvec{v'} \in \aut(w)$ to denote the fact that there is an accepting run of $\aut$ on $w$ whose cost is $\myvec{v'}$. 
Finally, we define the semantics of a CEFA $\aut$ as $\Lang(\aut) = \{(w; \myvec{v'}) \mid  \myvec{v'} \in \aut(w)\}$.

@@

%Finally, we define the product of two CEFA.

In the sequel, we define three operations on CEFA, namely, concatenation, intersection, and union. These three operations will be used for solving RECL constraints in the next section.

Let $\aut_1 = (R_1, Q_1, \Sigma, \delta_1, q_{1,0}, F_1, \alpha_1)$ and  $\aut_2 = (R_2, Q_2, \Sigma, \delta_2, q_{2,0}, F_2, \alpha_2)$ be two CEFA that share the alphabet. 
Moreover, suppose $R_1 \cap R_2 = \emptyset$. 
%We define two operations of CEFA, namely, the product and union, as follows.

The \emph{concatenation} of $\aut_1$ and $\aut_2$, denoted by $\aut_1 \concat \aut_2$, is

The \emph{product} of $\aut_1$ and $\aut_2$, denoted by $\aut_1 \times \aut_2 = (R', Q', \Sigma, \delta', q'_0, F', \alpha')$, is defined in the sequel. 
\begin{itemize}
\item $R' = R_1 \cup R_2$, $Q' = Q_1 \times Q_2$, $q'_0 = (q_{1,0}, q_{2,0})$, $F' = F_1 \times F_2$,
%
%\item $\alpha'((q_1, q_2)) = \alpha_1(q_1) \wedge \alpha_2(q_2)$ for each $(q_1, q_2) \in F'$, 
%
\item $\delta'$ comprises the tuples $((q_1, q_2), a, (q'_1, q'_2), \myvec{v})$ such that $(q_1,a, q'_1, \prj_{R_1}(\myvec{v})) \in \delta_1$ and $(q_2,a, q'_2, \prj_{R_2}(\myvec{v})) \in \delta_2$, where $\myvec{v} \in \Int^{R'}$, $\prj_{R_1}(\myvec{v}))$ (resp. $\prj_{R_2}(\myvec{v}))$) is the restriction of $\myvec{v}$ to the domain $R_1$ (resp. $R_2$). 
\end{itemize}
From the construction, $\Lang(\aut_1 \times \aut_2) = \{(w; \myvec{v'_1}, \myvec{v'_2}) \mid (w; \myvec{v'_1}) \in \Lang(\aut_1), (w; \myvec{v'_2}) \in \Lang(\aut_2)\}$. 
Intuitively, $\aut_1 \times \aut_2$ accepts the words that are accepted by both $\aut_1$ and $\aut_2$ and outputs the costs of $\aut_1$ and $\aut_2$.

The \emph{union} of $\aut_1$ and $\aut_2$

Let $((\aut_1,\cdots, \aut_n), \varphi)$ be a tuple such that $\aut_1 = (R_1, Q_1, \Sigma, \delta_1, q_{1,0}, F_1)$, $\cdots$, $\aut_n = (R_n, Q_n, \Sigma, \delta_n, q_{n, 0}, F_n)$ are CEFA, and $\varphi(R_1, \cdots, R_n)$ is an LIA formula. Then the \emph{relation} defined by $((\aut_1, \cdots, \aut_n), \varphi)$, denoted by $\srel((\aut_1, \cdots, \aut_n), \varphi)$, is $\{(w, \myvec{v'_1}, \cdots, \myvec{v'_n}) \mid \varphi(\myvec{v'_1}/R_1, \cdots, \myvec{v'_n}/R_n)\equiv \ltrue, \forall i \in [n].\ (w, \myvec{v'_i}) \in \Lang(\aut_i)\}$.

%the intersection of $\Lang(\aut_1)$ and $\Lang(\aut_2)$.

### 2023-05-03 2023-1-19/sections/preliminaries.tex
operations for NFA

\paragraph*{Finite automata.} 
A \emph{(nondeterministic) finite automaton} (NFA)  is a tuple $\NFA=(Q, \Sigma, \delta, q_0, F)$, where $Q$ is a finite set of states, $\Sigma$ is a finite alphabet, $\delta \subseteq Q \times \Sigma \times Q$ is the transition relation, $q_0 \in Q$ is the initial state, and $F \subseteq Q$ is the set of final states. For readability, we write a transition $(q, a, q') \in \delta$ as $q \xrightarrow[\delta]{a} q'$ (or simply $q \xrightarrow{a} q'$). %Moreover, when $\delta$ is clear from context, we omit $\delta$ in $q \xrightarrow[\delta]{a} q'$ and write $q \xrightarrow{a} q'$. 
The \emph{size} of an NFA $\NFA$, denoted by $|\NFA|$, is defined as the number of transitions of $\NFA$.
%
A \emph{run} of $\NFA$ on a string $w = a_1 \cdots a_n$ is a sequence of transitions $q_0 \xrightarrow{a_1} q_1 \cdots q_{n-1} \xrightarrow{a_n} q_n$. The run is \emph{accepting} if $q_n \in F$.
A string $w$ is accepted by an NFA $\NFA$ if there is an accepting run of $\NFA$ on $w$. In particular, the empty string $\varepsilon$ is accepted by $\NFA$ if $I \cap F \neq \emptyset$. The language of $\NFA$, denoted by $\Lang(\NFA)$, is the set of strings accepted by $\NFA$. 
%
An NFA $\NFA$ is said to be \emph{deterministic} if $I$ is a singleton and, for every $q \in Q$ and $a \in \Sigma$, there is at most one state $q' \in Q$ such that $(q, a, q') \in \delta$.
%
It is well-known that finite automata capture regular languages precisely. Moreover, the class of regular languages is closed under union, intersection, Kleene star, complement, and language difference \cite{HU79}.

>>>>

\paragraph*{Finite automata.} 
A \emph{(nondeterministic) finite automaton} (NFA)  is a tuple $\NFA=(Q, \Sigma, \delta, q_0, F)$, where $Q$ is a finite set of states, $\Sigma$ is a finite alphabet, $\delta \subseteq Q \times \Sigma \times Q$ is the transition relation, $q_0 \in Q$ is the initial state, and $F \subseteq Q$ is the set of final states. For readability, we write a transition $(q, a, q') \in \delta$ as $q \xrightarrow[\delta]{a} q'$ (or simply $q \xrightarrow{a} q'$). %Moreover, when $\delta$ is clear from context, we omit $\delta$ in $q \xrightarrow[\delta]{a} q'$ and write $q \xrightarrow{a} q'$. 
The \emph{size} of an NFA $\NFA$, denoted by $|\NFA|$, is defined as the number of transitions of $\NFA$.
%
A \emph{run} of $\NFA$ on a string $w = a_1 \cdots a_n$ is a sequence of transitions $q_0 \xrightarrow{a_1} q_1 \cdots q_{n-1} \xrightarrow{a_n} q_n$. The run is \emph{accepting} if $q_n \in F$.
A string $w$ is accepted by an NFA $\NFA$ if there is an accepting run of $\NFA$ on $w$. In particular, the empty string $\varepsilon$ is accepted by $\NFA$ if $I \cap F \neq \emptyset$. The language of $\NFA$, denoted by $\Lang(\NFA)$, is the set of strings accepted by $\NFA$. 
%
An NFA $\NFA$ is said to be \emph{deterministic} if $I$ is a singleton and, for every $q \in Q$ and $a \in \Sigma$, there is at most one state $q' \in Q$ such that $(q, a, q') \in \delta$. We use DFA to denote deterministic finite automata. 
%
It is well-known that finite automata capture regular languages. 
Moreover, the class of regular languages is closed under union, intersection, concatenation, Kleene star, complement, and language difference \cite{HU79}.  In the sequel, we recall the constructions of NFA corresponding to these language operations. Let $\NFA_1 = (Q_1, \Sigma, \delta_1, q_{1,0}, F_1)$ and $\NFA_2 = (Q_2, \Sigma, \delta_2, q_{2,0}, F_2)$ be two NFA such that $Q_1 \cap Q_2 = \emptyset$. 
\begin{itemize}
\item The union of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \cup \NFA_2$, is defined as the tuple $(Q', \Sigma, \delta', q'_0, F')$, where $Q'= Q_1 \cup Q_2 \cup \{q'_0\}$ with $q'_0 \not \in Q_1 \cup Q_2$, $\delta' = \delta_1 \cup \delta_2 \cup \{(q'_0, a, q') \mid (q_{1,0}, a, q') \in \delta_1\} \cup \{(q'_0, a, q') \mid (q_{2,0}, a, q') \in \delta_2\}$, $F'=F_1 \cup F_2$. 
%
\item The intersection of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \cap \NFA_2$, is defined as the tuple $(Q', \Sigma, \delta', q'_0, F')$, where $Q' = Q_1 \times Q_2$, $q'_0 = (q_{1,0}, q_{2,0})$, $F' = F_1 \times F_2$, $\delta' = \{((q_1, q_2), a, (q'_1, q'_2)) \mid (q_1, a, q'_1) \in \delta_1, (q_2, a, q'_2) \in \delta_2\}$.
%
\item The concatenation of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \concat \NFA_2$, is defined as the tuple $(Q', \Sigma, \delta', q'_0, F')$, where $Q' = Q_1 \cup Q_2$, $q'_0 = q_{1,0}$, $\delta' = \delta_1 \cup \delta_2 \cup \{(q_1, a, q_2) \mid q_1 \in F_1, (q_{2,0}, a, q_2) \in \delta_2\}$, moreover, if $q_{2,0} \in F_2$, then $F'= F_1 \cup F_2$, otherwise, $F'= F_2$.
%
\item The Kleene star of $\NFA_1$, denoted by $\NFA_1^*$, is defined as $(Q_1, \Sigma, \delta'_1, q_{1,0}, F'_1)$, where $\delta'_1 = \delta_1 \cup \{(q_1, a, q'_1) \mid q_1 \in F_1, (q_{1,0}, a, q'_1) \in \delta_1\}$, $F'_1= F_1 \cup \{q_{1,0}\}$.
%
\item The complement of $\NFA_1$, denoted by $\overline{\NFA_1}$, is defined as $(Q'_1, \Sigma, \delta'_1, \{q_{1,0}\}, Q'_1 \setminus F'_1)$, where  $(Q'_1, \Sigma, \delta'_1, \{q_{1,0}\}, F'_1)$ is the DFA obtained from $\NFA_1$ by subset construction.  
%
\item The language difference of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \setminus \NFA_2$, is defined as $\NFA_1 \times \overline{\NFA_2}$.
\end{itemize}
